Causal Inference in Population Trends: A Technique for finding Demographic Anomalies in Big Data

### 2020-01-20 MANUSCRIPT/manuscript.Rmd
Update manuscript.Rmd
Change to abstract and title

---
title: "Effects without a Cause: The Search for Demographic Anomalies"

@@

keywords:
- mortality
- fertility
- causal inference

abstract: |
  The proliferation of data, modern computing advances, and powerful statistical algorithms unlock the potential search for hidden, previously un- or understudied demographic anomalies. Here, we demonstrate how demographers can use causal inference techniques to identify anamolies by investigating US state-level fertility and mortality time series since 1999 to uncover the hidden baby booms/busts and mortality plagues?/non-plauges?. We find 38 states exhibited at least one mortality anomaly, totalling more than 318k anomalous deaths and an additional 6 states exhibited at least one mortality non-plague? totalling 164k protective deaths. 12 states exhibited baby busts, totalling more than 240k missing births and an additional 11 states exhibited baby booms totalling more than 134k additional births. These results suggest the widespread detection of demographic anomalies. Our analysis does not examine the *causes* of these anomalies and our results point to important further research on the causes of anomalous demographic behavior.

>>>>

---
title: "Causal Inference in Population Trends: A Technique for finding Demographic Anomalies in Big Data"

@@

keywords:
- mortality
- fertility
- causal inference
- abductive reasoning
- big data

abstract: |
The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in demographic data that have previously been dismissed as noise. Here, we demonstrate how demographers can use causal inference techniques and abductive reasoning to identify fertility and mortality anamolies in state-level data trends by investigating demographic time series data since 1999. In demonstrating our technique, we uncover real spikes and dips in fertility and mortality over time, distinguished from regular trend variations. We show that 38 states exhibited at least one mortality anomaly over time, totalling more than 318,000 additional deaths, while an additional six states exhibited periods of decline resuling in 164,000 fewer deaths. Twelve states exhibited real, marked periods of declines in fertility, totalling more than 240,000 fewer births than expected, and eleven states had periods of excess births, resulting in an addition 134,000 births. Our results show how researchers can detect demographic anomalies in trend data and use those anomalies as a beginning point for further investigation into the causes.
